Experience Report: Haskell in Computational Biology

### 2012-07-02 mrfy_experience_report.tex
looseness for later page breaking; edits pp 5-7

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took 50~lines of code and one~day to
implement.  
Simulated annealing required only a new
\texttt{utility} function, which took 15~lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm uses very similar
functions, except for \texttt{nextPt}:
recombination of parent
placements took forty lines of code
and a full day to implement.

@@

Even with this trick,
MRFy's implementation of Viterbi's algorithm is much
slower than the C++ version in \mrfy's predecessor,
SMURF.
For example, on a microbenchmark that searches for a
structural motif of 343~nodes in a protein of 2000 amino acids
using only Viterbi's algorithm and no beta-strand information,
\mrfy\ takes 2.32~seconds and
SMURF takes 0.29~seconds.

@@

We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
The model has 343~nodes, of which 178 appear in 40~beta strands.
The segments between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a lot of parallelism,
which Haskell made it easy to exploit.
Using
\texttt{Control.Parallel}, parallelizing the computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
\figref{fig:speedup} shows speedups when using from 
 1~to~48 of the cores 
on a 48-core, 2.3GHz AMD Opteron 6176 system.
After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
code, speedup rolls off.
By~running 4~instance of \mrfy\ in parallel on different searches,
we~hope to be able to use all 48~cores with about 50\%~efficiency.

@@

Our difficulties with debugging led to internal disagreements.
The~junior members of our team wanted to apply the debugging
skills they had honed through years of imperative programming.
These skills did not transfer well to Haskell.
The senior member of the team kept saying that any proper approach to
debugging Haskell code should involve QuickCheck.
This exhortation was unhelpful.

@@

Only the senior member of our team was able to use
QuickCheck easily.
In~retrospect, we~have identified some obstacles that prevented the
junior people from using
QuickCheck.
\begin{itemize}
\item
The examples and tutorials we found focused 
predominantly on writing and testing properties using data types that
already implemented class \texttt{Arbitrary}.
We~didn't understand the \texttt{Arbitrary} class very well, 
perhaps because the overloaded \texttt{arbitrary} value is not a
function.
For programmers accustomed to object-oriented dispatch
on \emph{arguments}, it is hard to grasp how Haskell's type-class
system can find the proper version
of \texttt{arbitrary}.
\item
Our difficulties were compounded by a weak understanding of monads.
We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
importance.
\item
We~continually overlooked the critical role of \emph{shrinking}.
As~a result, on the one or two occasions we did 
use QuickCheck, the counterexamples were too large to be
informative. 
\end{itemize}
Because of these obstacles, 
we wrote thousands of lines of code without ever defining an instance
of \texttt{Arbitrary} and therefore
without looking hard at QuickCheck.
After the fact, we were overwhelmed by the work
involved in writing and testing instances
of
\texttt{Arbitrary}.
The work got done only because the whole team got involved in order to
meet the deadlines for this paper.

Based on the QuickCheck properties we did write, we have identified and
easily fixed a bug in our implementation of Viterbi's algorithm; we had not
been including the score for the transition from the final node of the hidden
Markov model to the special ``end'' state.
Without QuickCheck, it is unlikely that we would have found this bug,
and if we had, determining the cause would have required much time.

@@

Our Haskell code for hidden Markov models and Viterbi's algorithm
solves the same problems as existing C++~code.
Other researchers using Haskell may also have to reimplement code,
but researchers may also reimplement code out of choice, not
necessity.
In computational biology, reimplementing existing algorithms is unremarkable.
For~example, both SMURF and HMMER contain new implementations of these
same algorithms.
In~MRFy, we are using the Haskell implementations to help solve new
problems.

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 2300 lines of~Haskell, 300 of which are devoted
to tests and QuickCheck properties and generators.
Neither C++ codebase includes any testing facilities.
MRFy~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
Its initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three of these tools were written
 in~haste.
We~have experience modifying the older tools.

@@

If~you are a computational biologist, 
you can be productive in Haskell without extensive preparation.
Two~of us (Daniels and Gallant) are graduate students.
Daniels has taken a graduate seminar in functional programming, which
included some Haskell;
Gallant has taken a programming-languages course which included
significant functional programming but no Haskell.
Ramsey is a professor who has used Haskell 
for medium-sized projects,
but his contributions to MRFy have been limited, mostly to \emph{post hoc}
refactoring and testing.

@@

While the Haskell community offers many
enticing tools, libraries, and packages,
be aware that
not all of them are worth using.
Some are not ready for prime time, and some were once great but are
no~longer maintained.
The great packages, like \texttt{Data.MemoCombinators} and Parallel
Strategies, are truly great.
But~for amateurs, it's not always easy to tell the great packages from the wannabes
and the has-beens.
And even some of the great packages could be better documented, with
more examples.

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.
ND and AG were funded in part by NIH grant 1R01GM080330 (to Lenore Cowen).

>>>>

Despite these non-modular aspects, the search
interface provides ample scope for
experiments.
Random hill climbing took 50~lines of code and one~day to
implement.  
Simulated annealing required only a new
\texttt{utility} function, which took 15~lines of code and half an hour to implement.
(Hill climbing accepts a point if and only if it scores better than its
predecessor;
simulated annealing may accept a point that scores worse.)
Our genetic algorithm uses very similar
functions, except for \texttt{nextPt}:
recombination of parent
placements took forty lines of code
and a full day to implement.
{\ifpagetuning\looseness=1 \par\fi}

@@

Even with this trick,
MRFy's implementation of Viterbi's algorithm is much
slower than the C++ version in \mrfy's predecessor,
SMURF.
\ifpagetuning\vadjust{\break}\fi
For example, on a microbenchmark that searches for a
structural motif of 343~nodes in a protein of 2000 amino acids
using only Viterbi's algorithm and no beta-strand information,
\mrfy\ takes 2.32~seconds and
SMURF takes 0.29~seconds.

@@

We also benchmarked \mrfy\ using a model of an ``8-bladed beta propeller.''
The model has 343~nodes, of which 178 appear in 40~beta strands.
The segments between beta strands typically have at most 10~nodes.
We~used a query sequence of 592 amino acids, but each placement breaks
the sequence into 41~pieces, each of which typically has at most 20 amino
acids.
Because \mrfy\ can solve the models between the beta strands independently,
this benchmark has a lot of parallelism,
which Haskell made it easy to exploit.
Using
\texttt{Control.Parallel}, parallelizing the computation was as easy
as  substituting
\texttt{parmap rseq} for \texttt{map}.
\figref{fig:speedup} shows speedups when using from 
 1~to~48 of the cores 
on a 48-core, 2.3GHz AMD Opteron 6176 system.
After about 12~cores, where \mrfy\ runs 6~times as fast as sequential
code, speedup rolls off.
By~running 4~instance of \mrfy\ in parallel on different searches,
we~hope to be able to use all 48~cores with about 50\%~efficiency.
{\ifpagetuning\looseness=1 \par\fi}

@@

Our difficulties with debugging led to internal disagreements.
The~junior members of our team wanted to apply the debugging
skills they had honed through years of imperative programming.
But these skills did not transfer well to Haskell.
The senior member of the team kept repeating that a proper approach to
debugging Haskell code should involve QuickCheck.
But mere exhortation was unhelpful.

@@

Only the senior member of our team was able to use
QuickCheck easily.
In~retrospect, we~have identified some obstacles that prevented the
junior people from using
QuickCheck.
\begin{itemize}
\item
The examples and tutorials we found focused 
predominantly on writing and testing properties using data types that
already implemented class \texttt{Arbitrary}.
We~didn't understand the \texttt{Arbitrary} class very well, 
perhaps because the overloaded \texttt{arbitrary} value is not a
function.
For programmers accustomed to object-oriented dispatch
on \emph{arguments}, it is hard to grasp how Haskell's type-class
system can find the proper version
of \texttt{arbitrary}.
\item
Our difficulties were compounded by a weak understanding of monads.
We~were too baffled by QuickCheck's \texttt{Gen} monad to grasp its
importance.
\item
We~continually overlooked the critical role of \emph{shrinking}.
As~a result, on the one or two occasions we did 
use QuickCheck, the counterexamples were too large to be
informative. 
\end{itemize}
Because of these obstacles, 
we wrote thousands of lines of code without ever defining an instance
of \texttt{Arbitrary} and therefore
without looking hard at QuickCheck.
After the fact, we were overwhelmed by the work
involved in writing and testing instances
of
\texttt{Arbitrary}.
The work got done only when the whole team pitched in to
meet the deadlines for this paper.

At the last minute, QuickCheck did find
a bug in our implementation of Viterbi's algorithm: we had omitted
the score for the transition from the final node of the hidden
Markov model to the special ``end'' state.
Without QuickCheck, we~probably wouldn't have known anything was wrong.
%it is unlikely that we would have found this bug,
%and if we had, determining the cause would have required much time.

@@

Our Haskell code for hidden Markov models and Viterbi's algorithm
solves the same problems as existing C++~code.
Other researchers using Haskell may also have to reimplement code,
but
in computational biology, reimplementing existing algorithms is unremarkable.
For~example, both SMURF and HMMER contain new implementations of
hidden Markov models and Viterbi's algorithm.

When performance has mattered, members of our group, like other
computational biologists, have used~C++.
To~compare our Haskell experience with our C++ experience,
we discuss three tools:
\begin{itemize}
\item
Matt~\cite{Menke:2008wu} is used to create alignments like that shown
in \figref{alignment}.
It~comprises about 12,000 lines of~C++.
The only external tools or libraries it uses are \texttt{zlib}
and \texttt{OpenMP}, and its initial development took two years.
\item
SMURF
\cite{Menke:2010ti} is used to detect homologous proteins in the presence
of paired beta strands.
It~comprises about 9,000 lines of~C++, of which about 1,000~lines are
shared with~Matt.
It~uses no external tools or libraries, and
its initial development took a year and a half.
It uses multidimensional dynamic programming to exactly compute the alignments
for which MRFy relies on stochastic search.
As a result, in the presence of complex beta-strand topologies, SMURF is
computationally intractable.
\item
MRFy is used to detect homologous proteins in the presence of paired
beta strands; it effectively supplants SMURF.
It~comprises about 2,500 lines of~Haskell, about 500 of which are devoted
to tests, QuickCheck properties, and generators.
Neither C++ codebase includes any test code.
MRFy~uses several external tools and libraries, of which the most
notable are Parsec, the BioHaskell bioinformatics library, and
the libraries \path{Data.MemoCombinators}, \path{Control.Parallel},
and \path{Data.Vector}.
\mrfy's initial development took about three months.
%%%% \remark{Others were mostly full-time work, but no need to distinguish.}
\end{itemize}
Like much research software, all three tools were written
 in~haste.
We~have experience modifying the older tools.

@@

If~you are a computational biologist and you are interested in
functional programming, you don't need extensive preparation
to be productive in Haskell.
Two~of us (Daniels and Gallant) are graduate students.
Daniels has taken a seminar in functional programming, which
included some Haskell;
Gallant has taken a programming-languages course which included
significant functional programming but no Haskell.
Ramsey is a professor who has used Haskell 
for medium-sized projects,
but his contributions to MRFy have been limited, mostly to \emph{post hoc}
refactoring and testing.

@@

While the Haskell community offers many
enticing tools, libraries, and packages,
not all of them are worth using.
Some are not ready for prime time, and some were once great but are
no~longer maintained.
The great packages, like \texttt{Data.MemoCombinators} and Parallel
Strategies, are truly great.
But~for amateurs, it's not always easy to tell the great packages from the wannabes
and the has-beens.
And even some of the great packages could be better documented, with
more examples.

@@

Anonymous referees spurred us to think
more deeply about laziness and to write more carefully about performance.
Josef Svenningsson suggested a correspondence between \mrfy's search
and stream fusion, which led us eventually to the \texttt{Utility} type.
Koen Claessen instantly diagnosed an inappropriately strict
implementation of the \texttt{Rand} monad.
We also thank Lenore Cowen, Kathleen Fisher, Ben Hescott, Brad
Larsen, and Nathan Ricci.

This work was funded in part by NIH grant 1R01GM080330.
